RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-08-31 doc/bioinformatics2013/RVD2.tex
Paper draft. Major update is on the goodness of fit test.

Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample. In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done 
\begin{algorithm}
\caption{Posterior distribution difference sampling}
\begin{algorithmic}[1]
\Procedure{}{}
\Repeat
\State  Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
\State  Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
\State Subtract the former from the later to get one sample for the posterior difference distribution 
\Until {Sample size is sufficient}
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

$p \left( \mu_{testing}-\mu_{ref} |r,n; \mu_0,M_0\right)$

\subsubsection{Bayesian Hypothesis Testing}
To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{testing}-\mu_{ref}\geq\tau \\
 H_2: \mu_{testing}-\mu_{ref}<\tau
\end{align}

@@

\begin{equation}
 p(H1:\mu_{testing}-\mu_{ref}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
\end{equation}

@@

\subsubsection{Uniformity Testing}

>>>>

Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample. In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done

\begin{enumerate}
 \item Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
 \item Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
 \item Subtract the former from the later to get one sample for the posterior difference distribution 
\end{enumerate}
Repeat the above three steps until sample size is sufficient.

@@

\subsubsection{Bayesian Hypothesis Testing}
To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{case}-\mu_{control}\geq\tau \\
 H_2: \mu_{case}-\mu_{control}<\tau
\end{align}

@@

\begin{equation}
 p(H1:\mu_{case}-\mu_{control}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
\end{equation}

@@

\subsubsection{Goodness of fit Testing for Uniformity}\

We assume that at one position in one experimental replicate, the non-reference reads follows multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.

The non-reference reads can be because of either mutation or various errors including sampling error, inter-replicated error, $etc$. In theory, the non-reference read counts caused by non-oriented errors should be uniformly distributed among three non-reference categories, namely $p_1=p_2=p_3=1/3$. However, the distribution of non-reference read counts among three non-reference categories caused by mutation theoretically is biased.

The null hypothesis can be denoted as
\begin{align}
 H_0: p=p_0
\end{align}
where
\begin{align} 
 p_0=(p_1,p_2,p_3), p_1=p_2=p_3=1/3
\end{align}

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984). Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

Person's $X^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family. Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.

Particulaly in our case, Let $(r_1,r_2,r_3)$ denotes the non-reference base frequencies and $\sum_{i=1}^3 r_i = r$. The expected frquencies are $(rp_1, rp_2, rp_3)$. Then the power-divergence statistics can be written as

\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^3 r_i \left[\left(\frac{r_i}{rp_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

We use $\lambda =2/3$ for the goodness of fit testing.

For each position we have six power-divergence statistics from 6 experimental replicates. We use Fisher's method to combines the six statistics into a single one for inference. Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.

Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.
